Crop adaptation through gene flow

### 2015-12-09 crop_adaptation_gene_flow_gj_outline.tex
updating crop_adaptation_gene_flow_gj_outline.tex, not quite done yet

\cite{bitocchi2009introgression} has a good review of the history of domestication, introduction to Europe, and further development.

\cite{hammer1987collection}\ cites an example in Italy of one domesticated maize adapting to coastal environment by introgressions from another domesticated maize.

@@

The first domesticated wheat was einkorn wheat (Triticum monococcum), domesticated from wild einkorn (T. m. boeoticum) in or around the Karacadag mountains of southeast Turkey \cite{heun1997site, harlan1966distribution, ozkan2002aflp}\.
Subsequently, emmer wheat (Triticum dicoccum) was domesticated from hybrids of diverse populations of wild emmer wheat (T. dioccoides) in the Fertile Crescent \cite{lev2000cradle, civavn2013reticulated, luo2007structure}.
Emmer wheat is tetraploid, and hybridization with wild diploid Tausch's goatgrass (Aegilops tauschii) lead to the production of hexaploid wheat \cite{salamini2002genetics, hancock2012plant, dvorak2006molecular}\.
Emmer wheat has high genetic diversity, likely due to gene flow from wild emmer \cite{luo2007structure, dvorak2006molecular}\.

Durum wheat (T. turgidum ssp. durum) was developed from emmer wheat as well.

Common bread wheat, T. aestivum, is an allohexaploid with chromosome sets from red wild einkorn wheat (T. urartu), Aegilops speltoides, and Aegilops tauschii \cite{mcfadden1946origin, petersen2006phylogenetic}\, and furthermore, each of these species is the result of ancient hybridizations \cite{marcussen2014ancient}\.
Hexaploidization probably occured twice, in southeast Turkey/northern Syria and, more recently, Iran \cite{giles2006gludy}\.

Domesticated einkorn wheat (Triticum monococcum monococcum), wild einkorn wheat (T. m. boeoticum), and the progenitor T. m. boeoticum are from the Karacadag [[special character]] mountains of southeast Turkey \cite{heun1997site}.
Wild emmer wheat (Triticum turgidum subsp. dicoccoides), the progenitor of modern (tetraploid and hexaploid) emmer wheat, is from the Fertile Crescent \cite{lev2000cradle, ozkan2002aflp}.
Two distinct taxa of Triticum dioccoides exist (a western one and a central-eastern one), but the central-eastern one was the progenitor species from which modern wheat was domesticated \cite{ozkan2005reconsideration}.
Localizing domestication centers for other forms of wheat at a resolution greater than the Fertile Crescent is not yet possible \cite{nesbitt1998wheat}, but then \cite{lev2000cradle} point to "a small core area within the Fertile Crescent—near the upper reaches of the Tigris and Euphrates rivers [HN6] in present-day southeastern Turkey/northern Syria".
So, wheat has one restricted center of domestication, with wild relatives nearby.
From \cite{ozkan2005reconsideration}\, "The geographical distribution of T. dicoccoides reported by Zohary and Hopf (2000) includes the western Fertile Crescent, the central part of southeastern Turkey and areas in eastern Iran and Iraq. Johnson (1975) reports that from southeastern Turkey to Iran and Iraq the species is progressively substituted by the wild tetraploid wheat T. araraticum. T. dicoccoides also grows on the basaltic rocky slopes of the Karacadag mountains in southeastern Turkey (Johnson 1975; Harlan and Zohary 1966).

The progressive domestication of wheat has involved many introgressive acts between geographically separated lineages, evinced by identification of genomes' various sources in wheats with elevated ploidy levels [[see the two papers i included for bread wheat background]]

The a, b, c, and f alleles of mildew resistance gene Pm3 may have been introgressed into wheat from wild emmer wheat shortly after domestication (although the e, d, and g alleles were probably formed by de novo mutations) \cite{TPJ:TPJ2772}.
Stem rust resistance bred into cultivated T. durum (tetraploid) from wild einkorn wheat \cite{gerechter1971transfer}\.
Pm36, a novel podery mildew resistance gene, was also experimentally introgressed from T. turgidum var. dicoccoides into durum wheat \cite{blanco2008molecular}\.

@@

Domesticated and wild barleys belong to the same species, Hordeum vulgare, and are biologically capable of producing viable offpsring via hybridizaion \cite{von1995ecographical}.
Barley (Hordeum vulgare subsp. vulgare) is believed to have been domesticated from wild subsp. spontaneum in the Fertile Crescent and by subsp. spontaneum var. agriocrithon in Tibet roughly 10,000 years ago \cite{takahashi1955origin, badr2000origin, oka2012origin, azhaguvel2007phylogenetic, haberer2015barley}\, but many of the details of barley domestication are still disputed.

Presently, the distribution of spontaneum consists of the Mediterranean, the Middle-East, and West-Central Asia, while other barley wild relatives inhabit Asiatic regions (including Tibet) more broadly \cite{nevo2010drought, harlan1995living}\.
These wild relatives inhabit regions spanning such abiotic clines as temperature, precipitation, soil type, and altitude, as well as biotic clines \cite{nevo2010drought}\.

The barley domestication process has reduced the number of alleles in the domesticate to only 40% of that found in wild barley, though there remains a great deal of phenotypic diversity among the wild barleys \cite{ellis2000wild}\.

Variety agriocrithon (H. vulgare ssp. vulgare f. agriocrithon) is genetically diverse, and is found throughout much of the range of Barley.
Some have suggested that agriocrithon may be the progenitor of six-rowed barley, the product of a hybridization between eastern and western cultivated barley, or a wild-domesticate hybrid \cite{staudt1961origin, zohary1959hordeum, murphy1982origin}\, but \cite{azhaguvel2007phylogenetic} dispute these theories, suggesting instead that it arose from ssp. spontaneum, perhaps more than once (in Israel as well as in Tibet).
Wild-domesticate breeding experiments have shown that wild barleys have alleles for several important agronomic phenotypes, including brittleness, flowering time, plant height, lodging, and yield,  \cite{von2006ab, handley1994chromosome}\.
See art%0223-4 for list of papers that discuss the beneficial effect of exotic genes on biotic, abiotic, and quality traits.

Oweing to the lack of evidence of adaptive introgression between wild and domesticated barley is the low rate of outcrossing of spontaneum, estimated by \cite{brown1978outcrossing} to be 1.6% in Israeli populations.

Barley/H. spontaneum hybrids are fertile, but hybrids of other wild relatives generally are not.
There has been little genetic investigation into spontaneous barley/spontaneum hybrids \cite{ellstrand2003dangerous}

Look, there's no barrier to hybridization, there was a broad suite of wild barleys already in place, they're all locally adapted, this is a prime example of POSSIBLE introgression, but where's the evidence?

Introgression from wild to domesticate has been shown to happen over distances of more than a kilometer \cite{hillman2001new}\.

@@

Helianthus annuus shows evidence of domestication in eastern United States \cite{harter2004origin, wills2006chloroplast}\, with additional evidence of a possible second origin of domestication in Mexico \cite{lentz2008sunflower}\.
Helianthus is a versatile species, capable of adapting to a wide range of environments across the Americas, Europe, Asia, and Australia \cite{kane2008genetics}\.
[[This versatility may be due in part either to phenotypic plasticity or genetic adaptability \cite{maron2004rapid}\.]] This sentence may not help much, actually.  We're betting on genetic adaptability via introg hyb

H. annus adapted to the environment in Texas by hybridizing with H. debilis ssp. cucumerifolius, gaining advantageous alleles (these hybrids are now called H. annus ssp. texanus, \cite{kim1999genetic, heiser1951hybridization, rieseberg1999hybrid, rieseberg1990helianthus}.
Helianthus has several genes for downy mildew resistance.
Each imparts resistance to one or more races of P. halstedii, one of the most agronomically important diseases in sunflower cultivation \cite{cohen1973factors}\.
Some of these downy mildew resistence genes were found in wild relatives (including H. argophyllus, H. tuberosus, and H. praecox) and have been successfully bred into modern H. annus \cite{miller1991inheritance}\.
PlArg, an allele found in wild silverleaf sunflowers (Helianthus argophyllus, inbred line Arg1575-2), confers resistance to all known (20 or more) races of downey mildew \cite{dussle2004pl}\, while others (Pl1-Pl11) are effective for one or more types \cite{rahim2002inheritance}\.
H. annus lives in sympatry with wild relatives like H. petiolaris and H. bolanderi and forms  stable hybrid populations \cite{schwarzbach2002likely, rieseberg1988molecular, welch2002patterns}\.
H. annus shows signs of persistent introgressive hybridization with H. petiolaris with evidence of positive selection driving some of the genetic differentiation between the two species \cite{yatabe2007rampant}\.
\cite{kane2008genetics} finds that most gene flow in weedy Helianthus is within geographic regions, rather than between separate popultions.
Wild sunflowers are locally adapted, and weedy hybrid populations share these adaptations \cite{kane2008genetics}\.
Weediness is a trait that has likely evolved many times \cite{kane2008genetics}\.

Silverleaf sunflower has also been the focus of drought resistance breeding efforts \cite{saucă2010introgression}\ and Phomopsis resistance breeding efforts \cite{besnard1997specifying}\.

@@

There are two main cultigens of rice; Asian rice, Oryza sativa, and African rice, O. glaberrima, which are easily distinguish by ligule length, number of secondary panicle branches, panicle axis thickness, and because O. glaberrima is annual \cite{oka2012origin}.

The centers of rice domestication are not known with complete confidence, but genetic and archaeobotanical evidence points towards both the Yangzee Basin in China and the Ganges plains in India for O. sativa, 8,200-13,500 years ago, from wild O. rufipogon  \cite{oka2012origin, fuller2010consilience, ricepedia, molina2011molecular}\ and the Upper Niger River delta in Mali, Africa for O. glaberrima, 2-3,000 years ago, from ancestor O. barthii \cite{ricepedia}\.

The two main subspecies of Asian rice, indica and japonica, were domesticated from wild O. rufipogon "or the Asian form of O. perennis complex" \cite{oka2012origin}\ or O. nivara, which is closely related to O. rufipogon \cite{sharma2000origin, oka1988origin}\, whereas O. glaberrima was domesticated from O. breviligulata \cite{oka2012origin}\.

Both O. sativa/O. rufipogon and O. glaberrima/O.breviligulata produce weedy hybrids.
The greater genetic diversity within O. sativa is likely due to introgression with wild relatives both during domestication and upon the dispersal of O. sativa into new environments and sympatry with new relatives \cite{second1982origin}.

O. sativa has two main subspecies, Indica (with subpopulations indica and aus) and Japonica (with subpopulations temperate japonica, tropical japonica known as javanica, and aromatic) \cite{chang2003origin, glaszmann1987isozymes, ricepedia}.
These subspecies' display adaptations to the environmental coniditions corresponding to the differentiated geographical locations they inhabit \cite{khush2003classifying}\.

There are about 20 other wild species within the genus Oryza \cite{ricepedia}\.

There exists evidence of gene flow between rice subpopulations and into wild rice \cite{zhao2010genomic}\.

Several resistance genes (grassy stunt virus, bacterial blight, brown planthopper, blast) have been introgressed from wild relatives into O. sativa L. by researchers \cite{brar1997alien, khush1974inheritance}, but adaptive introgression has also happened without human intervention \cite{second1982origin}.

@@

Rye (Secale cereale) was most likely domesticated from S. vavilovii \cite{stutz1972origin}\.
The species Secale cereale contains domesticated, weedy, and wild rye subspecies \cite{khush1961cytogenetic}\.
Rye domestication has recieved comparatively little academic interest \cite
N. I. Vavilov (the namesake of the species name of this plant) hypothesized that rye originated as a secondary crop during the domestications of wheat, barley, and/or peas \cite{vavilov1928geographical}\.

@@

Modern potato Solanum tuberosum is believed to have been domesticated in southern Peru in sympatry with a multitude of wild relatives about 6000 years ago, although the exact location and formal classification and phylogenetic relationships between these taxa have long been disputed \cite{huaman2002reclassification, spooner2005single, pickersgill1977origins, hawkes1988evolution}\.
The northern members of the polyphyletic S. brevicaule wild potato complex have been identified as likely progenitors of modern potatoes \cite{correll1962potato}, but determination of a single progenitor species is unlikely, either because widespread gene flow in the complex will mask signs of this phylogeny or because s. tuberosum has a polyphyletic origin.
Then again, these northern members of the S. breviaule complex are not clearly defined, and may in fact be one singular species (in which case the species name would be S. bukasovii) \cite{spooner2005single}.

>>>>

\cite{bitocchi2009introgression} has a good review of the history of domestication, introduction to Europe, and further development.
\cite{hammer1987collection}\ cites an example in Italy of one domesticated maize adapting to coastal environment by introgressions from another domesticated maize.

@@

The first domesticated wheat was einkorn wheat (Triticum monococcum), domesticated from wild einkorn (T. m. boeoticum) in or around the Karacadag mountains of southeast Turkey \cite{heun1997site, harlan1966distribution, ozkan2002aflp}\.
Subsequently, emmer wheat (Triticum dicoccum) was domesticated from hybrids of diverse populations of wild emmer wheat (Triticum dioccoides), also in the Fertile Crescent \cite{lev2000cradle, civavn2013reticulated, luo2007structure, ozkan2002aflp}.
Two distinct taxa of T. dioccoides exist (a western one and a central-eastern one), but the central-eastern one was the progenitor species from which modern wheat was domesticated \cite{ozkan2005reconsideration}
The tetraploid durum wheat (Triticum turgidum) is the result of hybridization between red wild einkorn wheat (T. urartu) and wild goatgrass (Aegilops speltoides) \cite{mcfadden1946origin, petersen2006phylogenetic}\, and a further hybridization with Tausch's goatgrass (Aegilops tauschii) produced the allohexaploid common bread wheat (T. aestivum), the most agronomically important race of wheat \cite{salamini2002genetics, hancock2012plant, dvorak2006molecular}\.
This hexaploidization event probably occured twice, once in southeast Turkey/northern Syria and, more recently, in Iran \cite{giles2006gludy}\.
Each of the three genomes that comprise bread wheat shows genomic signatures of ancestral hybridizations as well \cite{marcussen2014ancient}\.

The progressive domestication of wheat has involved many introgressive acts between geographically dispersed lineages, evinced by identification of genomes' various sources in wheats with elevated ploidy levels.

The convoluted domestication of wheat is beset by many instances of hybridization between morphologically-distinct species within a restrained geographical range.
Throughout wheat domestication, there has been gene flow between species and between wild and domesticated wheats.
Emmer wheat, for example, has high genetic diversity, likely due to gene flow from wild emmer \cite{luo2007structure, dvorak2006molecular}\.
The a, b, c, and f alleles of mildew resistance gene Pm3 may have been introgressed into wheat from wild emmer wheat shortly after domestication (although the e, d, and g alleles were probably formed by de novo mutations) \cite{TPJ:TPJ2772}.
Furthermore, wild wheat relatives are can be found in a much broader swath than the domesication center in the Fertile Crescent \cite{CWR}\.
Modern breeding efforts between wild and domesticated wheat have shown some success.
For example, some researchers have experimentally bred resistance traits into durum wheat (stem rust resistance from wild einkorn, \cite{gerechter1971transfer}\, and powdery mildew resistance from T. turgidum var. dicoccoides, \cite{blanco2008molecular}\.

@@

Domesticated and wild barleys belong to the same species, Hordeum vulgare, and are biologically capable of producing viable offpsring via hybridizaion \cite{von1995ecographical}.
Barley (Hordeum vulgare subsp. vulgare) is believed to have been domesticated at least twice, once from wild subsp. spontaneum in the Fertile Crescent and once by subsp. spontaneum var. agriocrithon in Tibet roughly 10,000 years ago \cite{takahashi1955origin, badr2000origin, oka2012origin, azhaguvel2007phylogenetic, haberer2015barley}\, but many of the details of barley domestication are still disputed.

Variety agriocrithon (H. vulgare ssp. vulgare f. agriocrithon) is genetically diverse, and is found throughout much of the range of Barley.
Some have suggested that agriocrithon may be the progenitor of six-rowed barley, the product of a hybridization between eastern and western cultivated barley, or a wild-domesticate hybrid \cite{staudt1961origin, zohary1959hordeum, murphy1982origin}\, but \cite{azhaguvel2007phylogenetic}\ dispute these theories, suggesting instead that it arose from ssp. spontaneum, perhaps more than once (in Israel as well as in Tibet).
Presently, the distribution of spontaneum consists of the Mediterranean, the Middle-East, and West-Central Asia, while other barley wild relatives inhabit Asiatic regions (including Tibet) more broadly \cite{nevo2010drought, harlan1995living, CWR}\.
These wild relatives inhabit regions spanning such abiotic clines as temperature, precipitation, soil type, and altitude, as well as biotic clines \cite{nevo2010drought}\.

The barley domestication process has reduced the number of alleles in the domesticate to only 40% of that found in wild barley, though there remains a great deal of phenotypic diversity among the wild barleys \cite{ellis2000wild}\.
Wild-domesticate breeding experiments have shown that wild barleys have alleles for several important agronomic phenotypes, including brittleness, flowering time, plant height, lodging, and yield,  \cite{von2006ab, handley1994chromosome}\.
Although the conditions of barley domestication would seem to allow (if not promote) natural adaptive introgression between barley and its wild relatives, there is little evidence of this genetic interaction at present.
Owing to this sparcity of evidence is the low rate of outcrossing of spontaneum, estimated by \cite{brown1978outcrossing} to be 1.6% in Israeli populations.
There has been little genetic investigation into spontaneous barley/spontaneum hybrids \cite{ellstrand2003dangerous}.
Barley/H. spontaneum hybrids are fertile, and morphologically intermediate (putatively hybrid) barleys are found when wild and domesticated barleys are grown in sympatry, but hybrids of other wild relatives generally exhibit greatly diminished fertility \cite{ellstrand2003dangerous, harlan1995living}\.
Even when the two are not grown immediately adjacent to one another, introgression from wild to domesticate has been shown to happen over distances of more than a kilometer \cite{hillman2001new}\.

@@

The common sunflower (Helianthus annuus) shows evidence of domestication in eastern United States \cite{harter2004origin, wills2006chloroplast}\, with additional evidence of a possible second origin of domestication in Mexico \cite{lentz2008sunflower}\.
Helianthus is a versatile species, capable of adapting to a wide range of environments across the Americas, Europe, Asia, and Australia \cite{kane2008genetics}\.
[[This versatility may be due in part either to phenotypic plasticity or genetic adaptability \cite{maron2004rapid}\.]] This sentence may not help much, actually.  We're betting on genetic adaptability via introg hyb
H. annuus adapted to the environment in Texas by hybridizing with H. debilis ssp. cucumerifolius, gaining advantageous alleles (these hybrids are now called H. annuus ssp. texanus, \cite{kim1999genetic, heiser1951hybridization, rieseberg1999hybrid, rieseberg1990helianthus}.
H. annuus lives in sympatry with wild relatives like H. petiolaris and H. bolanderi and forms  stable hybrid populations \cite{schwarzbach2002likely, rieseberg1988molecular, welch2002patterns}\.
H. annuus shows signs of persistent introgressive hybridization with H. petiolaris with evidence of positive selection driving some of the genetic differentiation between the two species \cite{yatabe2007rampant}\.
Wild sunflowers are locally adapted, and weedy hybrid populations share these adaptations \cite{kane2008genetics}\.
Weediness is a trait that has likely evolved many times \cite{kane2008genetics}\.

Helianthus has several genes for downy mildew resistance.
Each imparts resistance to one or more races of P. halstedii, one of the most agronomically important diseases in sunflower cultivation \cite{cohen1973factors}\.
Some of these downy mildew resistence genes were found in wild relatives (including H. argophyllus, H. tuberosus, and H. praecox) and have been successfully bred into modern H. annuus \cite{miller1991inheritance}\.
PlArg, an allele found in wild silverleaf sunflowers (Helianthus argophyllus, inbred line Arg1575-2), confers resistance to all known (20 or more) races of downey mildew \cite{dussle2004pl}\, while others (Pl1-Pl11) are effective for one or more types \cite{rahim2002inheritance}\.
Silverleaf sunflower has also been the focus of drought resistance breeding efforts \cite{saucă2010introgression}\ and Phomopsis resistance breeding efforts \cite{besnard1997specifying}\.

@@

There are two main cultigens of rice; Asian rice (Oryza sativa) and African rice (O. glaberrima), which are easily distinguish by ligule length, number of secondary panicle branches, panicle axis thickness, and differences in life cycles \cite{oka2012origin}.
The centers of rice domestication are not known with complete confidence, but genetic and archaeobotanical evidence points towards both the Yangzee Basin in China and the Ganges plains in India for O. sativa, 8,200-13,500 years ago, from wild O. rufipogon  \cite{oka2012origin, fuller2010consilience, ricepedia, molina2011molecular}\ or "the Asian form of O. perennis complex" \cite{oka2012origin}\ and the Upper Niger River delta in Mali, Africa for O. glaberrima, 2-3,000 years ago, from a wild ancestor, perhaps O. barthii \cite{ricepedia}\ or O. breviligulata \cite{oka2012origin}\.
Asian rice has two main subspecies, Indica (with subpopulations indica and aus) and Japonica (with subpopulations temperate japonica, tropical japonica known as javanica, and aromatic) \cite{chang2003origin, glaszmann1987isozymes, ricepedia}.
Indica and Japonica are likely the result of independent domestications from separate O. rufipogon populations in India/Indochina and southern China, respectively \cite{londo2006phylogeography}\.
These subspecies display adaptations to the environmental coniditions corresponding to the differentiated geographical locations they inhabit \cite{khush2003classifying}\.

Both Asian and African rice naturally hybridize with other domesticated subspecies and with wild relatives (of which there are about twenty, \cite{ricepedia}\), and introgression is common \cite{oka2012origin, second1982origin, zhao2010genomic}\.
The greater genetic diversity within O. sativa is likely due to introgression with wild relatives both during domestication and upon the dispersal of O. sativa into new environments and sympatry with new relatives \cite{second1982origin}.
Several resistance genes (grassy stunt virus, bacterial blight, brown planthopper, blast) have been introgressed from wild relatives into O. sativa L. by researchers \cite{brar1997alien, khush1974inheritance}.
Beyond investigative experiments, gene flow from wild relatives has been used to produce agronomic rice varieties.
Yatsen No. 1, for example, showed resistance to pests and diseases and adapted well to environmental conditions \cite{ting1933wild}\.
Several lines were derived from Yatsen No. 1, and went on to be utilized extensively in parts of China.

@@

Rye domestication has recieved comparatively little academic interest.
Although there is little evidence for or against the role of natural introgression in rye domestication, rye (Secale cereale) has wild relatives outside its center of domestication in or around the Fertile Crescent \cite{vavilov1928geographical}\, and the species contains domesticated, weedy, and wild rye subspecies \cite{khush1961cytogenetic}\.

@@

Modern potato (Solanum tuberosum) is believed to have been domesticated in southern Peru in sympatry with a multitude of wild relatives about 6000 years ago, although the exact location and formal classification and phylogenetic relationships between these taxa have long been disputed \cite{huaman2002reclassification, spooner2005single, pickersgill1977origins, hawkes1988evolution}\.
The northern members of the polyphyletic S. brevicaule wild potato complex have been identified as likely progenitors of modern potatoes \cite{correll1962potato}, but determination of a single progenitor species is unlikely, either because widespread gene flow in the complex will mask signs of this phylogeny or because s. tuberosum has a polyphyletic origin.
Also, these northern members of the S. breviaule complex are not clearly defined, and may in fact be one singular species (in which case the species name would be S. bukasovii, \cite{spooner2005single}\).
